{\small} High Performance P$^{3}$M N-body code: CUBEP3M:

### 2012-05-23 Extendedpp.tex
Introduced a naming of the simulation runs for easier reference

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.
 To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_ppext6} a force versus distance
 plot, analogous to Fig. \ref{fig:den_force_ppext0}, but the pp force has been extended to six layers of fine cells. 
 We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
 The same applies to the fractional error on the radial and tangential components of the force, as seen in Fig. \ref{fig:den_force_fracErr_ppext6},
 which are now at least 5 time smaller than in the default p$^{3}$m algorithm.

 \begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
  \caption{Gravity force in of the p$^3$m algorithm, compared with the exact $1/r^{2}$ law,
  with the pp force extended to six fine mesh layers.
  Particles in the same cell follow the exact curve, then we observe a large scatter at 
  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
    \label{fig:den_force_ppext6}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
  \caption{Fractional error on the force in of the p$^3$m algorithm, in the radial (top) and tangential(bottom) directions,
  with the pp force extended to six fine mesh layers.
    \label{fig:den_force_fracErr_ppext6}}
\end{center}
\end{figure}

 To quantify the accuracy improvement versus computing time requirements, we performed the following test.
 We generate a set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
 and with $128^{3}$ particles. We evolve the particles to $z=0$ with different ranges for the pp calculation, and compare 
 the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
 such that the only difference between different runs is the range of the pp force.
 Fig. \ref{fig:power} shows the dimensionless power spectrum of the different runs, where we see a significant gains in resolution
 when extending  PM to P$^{3}$M first, and when adding successive layers of fine cell mesh where the pp force is extended.
We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the fine grid calculations
are more accurate as the distance increases. For this reason, it seems that a range of a few layers, between 3 and 6, suffices 
to reduce most of the undesired NGP scatter.

@@

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.}
    \label{fig:power}
  \end{center}
\end{figure}

>>>>

 To speed up the access to all particles within the domain of computation, we construct a thread safe linked list
 to be constructed and accessed in parallel by each core of the system, but this time with a head-of-chain that points to the first particle in the current fine mesh cell. We then loop over all fine grids, accessing the particles contained therein and inside each fine grid cells for which we killed the mesh kernels,
 we compute the separation and the force between each pairs and update their velocities simultaneously with Newton's third law. 
 To avoid double counting, we loop only over the fine mesh neighbours that produce non-redundant contributions. Namely, for a central cell located at 
 $(x_1, y_1, z_1)$, we only consider the neighbours $(x_2, y_2, z_2)$ that satisfy the following conditions:
 \begin{itemize}
 \item{$z_2 \ge z_1$ always}
 \item{if $z_2 = z_1$, then $y_2 \ge y_1$, otherwise we also allow $y_2 < y_1$} 
 \item{if $z_2 = z_1$ and $y_2 = y_1$, then we enforce $x_2 > x_1$}
 \end{itemize}
 The case where all three coordinates are equal is already calculated in the standard configuration of the code.
 To assess the improvement of the force calculation, we present in Fig \ref{fig:den_force_ppext6} a force versus distance
 plot in a CITA256 realization , analogous to Fig. \ref{fig:den_force_ppext0}, but the pp force has been extended to six layers of fine cells. 
 We observe that the scatter about the theoretical curve has reduced significantly, and is still well balanced around the theoretical predictions.
 The same applies to the fractional error on the radial and tangential components of the force, as seen in Fig. \ref{fig:den_force_fracErr_ppext6},
 which are now at least 5 time smaller than in the default p$^{3}$m algorithm.

 \begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=6.eps}
  \caption{Gravity force in the p$^3$m algorithm, compared with the exact $1/r^{2}$ law,
  in a CITA256 realization. The exact pp force has been extended to six fine mesh layers around each particle.
  Particles in the that range follow the exact curve, then we observe a much smaller scatter at 
  distance of the order of 6 fine grid, compared to Fig. \ref{fig:den_force_ppext0}. 
  The NGP interpolation scheme is again responsible for the scatter, but the effect is suppressed with increasing distances. 
    \label{fig:den_force_ppext6}}
\end{center}
\end{figure}

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_fracErr_ppext=6.eps}
  \caption{Fractional error on the force in the p$^3$m algorithm, in the radial (top) and tangential (bottom) directions,
  in a CITA256 simulation in which the exact pp force extends to six fine mesh layers.
    \label{fig:den_force_fracErr_ppext6}}
\end{center}
\end{figure}

 To quantify the accuracy improvement versus computing time requirements, we performed the following test.
 We generate a set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$,
 and with $128^{3}$ particles. We evolve these CITA128 realizations with different ranges for the pp calculation, and compare 
 the resulting power spectra. For the results to be meaningful, we also need to use the same random seed for the random number generator,
 such that the only difference between different runs is the range of the pp force.
 Fig. \ref{fig:power} shows the dimensionless power spectrum of the different runs, where we see a significant gains in resolution
 when extending  PM to P$^{3}$M first, and when adding successive layers of fine cell mesh where the pp force is extended.
We have not plotted the results for higher numbers of layers, as the improvement becomes milder there: the fine grid calculations
are more accurate as the distance increases. For this reason, it seems that a range of a few layers, between 3 and 6, suffices 
to reduce most of the undesired NGP scatter.

@@

\begin{figure}
  \begin{center}
  \epsfig{file=graphs/pp_ext.eps,width=0.49\textwidth}
  \caption{ Dimensionless power spectrum for varying range of the exact pp force, compared to  {\small CAMB} \citep{Lewis:1999bs}.
  These CITA128 realizations evolved from a unique set of initial conditions at a starting redshift of $z = 100$, with a box size equal to $ 200 h^{-1}\mbox{Mpc}$ 
  until $z=0$. The only difference between the runs is the ranges of the pp calculation.}
    \label{fig:power}
  \end{center}
\end{figure}

### 2012-05-23 Halos.tex
Introduced a naming of the simulation runs for easier reference

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
  \end{center}
  \caption{Simulated halo multiplicity function, 
    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
    simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
    particles (red, solid). For reference we also show a widely-used 
    precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
    \label{mf}}
\end{figure}

A sample halo mass function produced based on our inlined SO halo finder 
at redshift $z=0$ from a simulation with $3.2\,h^{-1}Gpc$ box and $4000^3$
particles is shown in Fig. \ref{mf}. We compare our result to the 
precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
other widely-used fits like the one by \citet{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
halo finders, the \citet{2008ApJ...688..709T} fit is based on the
SO search algorithm, whose masses are systematically different 
from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
making this fit a better base for comparison here. Results show excellent
agreement, within $\sim10$ per cent for all halos with masses corresponding to
1000 particles or more. Lower-mass halos are somewhat under-counted compared
to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
SO halo finder, which misses some of the low-mass halos. Using more sophisticated
halo finders (available through post-processing due to their heavier memory
footprint) recovers the expected mass function.

>>>>

\begin{figure}%[ht]
%  \vskip -0.5cm 
  \begin{center}
    \includegraphics[width=3.2in]{graphs/mf_z0_Tinker.eps}
  \end{center}
  \caption{Simulated halo multiplicity function, 
    $\frac{M^2}{\bar{\rho}}\frac{dn}{dM}$ based on a
    RANGER4000 simulation with $3.2\,h^{-1} \mbox{Gpc}$ box and $4000^3$ 
    particles (red, solid). For reference we also show a widely-used 
    precise fit by \citet{2008ApJ...688..709T} (blue, dashed). 
    \label{mf}}
\end{figure}

A sample halo mass function produced based on our inlined SO halo finder 
at redshift $z=0$ from a RANGER4000 simulation is shown in Fig. \ref{mf}. We compare our result to the 
precise fit presented recently by \citet{2008ApJ...688..709T}. Unlike most
other widely-used fits like the one by \citet{2002MNRAS.329...61S}, which are based on friends-of-friends (FOF)
halo finders, the \citet{2008ApJ...688..709T} fit is based on the
SO search algorithm, whose masses are systematically different 
from the FOF masses \citep[e.g.][]{2007MNRAS.374....2R,2008ApJ...688..709T}, 
making this fit a better base for comparison here. Results show excellent
agreement, within $\sim10$ per cent for all halos with masses corresponding to
1000 particles or more. Lower-mass halos are somewhat under-counted compared
to the \citet{2008ApJ...688..709T} fit, by $\sim20$ per cent for 400 particles and 
by $\sim40$ per cent for 50 particles. This is due to the grid-based nature of our
SO halo finder, which misses some of the low-mass halos. Using more sophisticated
halo finders (available through post-processing due to their heavier memory
footprint) recovers the expected mass function.

### 2012-05-23 Initial.tex
Introduced a naming of the simulation runs for easier reference

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
Fig. \ref{fig:init} shows the late time power spectrum of two realizations that started off the same initial power spectrum, 
but one of which had non-Gaussian initial conditions.
We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
agrees much better with one loop perturbation theory \citep{XXX}.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50_w_1loop_PT.eps}
  \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The two curves represent two realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator. The to curves differ at the sub-per cent level, as seen in the top panel.   
    \label{fig:init}}
\end{center}
\end{figure}

>>>>

This code was tested by comparing simulations and theoretical predictions
for the effect of local primordial non-Gaussianity on the halo mass 
function and matter power spectrum (Desjacques, Seljak \& Iliev 2009). 
It has also been used to quantify the impact of local non-Gaussian initial
conditions on the halo power spectrum \citep{2009MNRAS.396...85D,
2010PhRvD..81b3006D} and bispectrum \citep{2010MNRAS.406.1014S},
 as well as the matter bispectrum \citep{2011arXiv1111.6966S}.
Fig. \ref{fig:init} shows the late time power spectrum of two RANGER4000 realizations that started off the same initial power spectrum, 
but one of which had non-Gaussian initial conditions.
We see that the difference between the two power spectra is at the sub-per cent level, but that the non-Gaussian realization
agrees much better with one loop perturbation theory \citep{XXX}.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/dm_power_z0_fNL0_50_w_1loop_PT.eps}
  \caption{ Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The two curves represent two RANGER4000 realizations of the same initial power spectrum, one of which used Gaussian statistics (blue) and the other the non-Gaussian initial condition generator. The two curves differ at the sub-per cent level, as seen in the top panel,
  and the non-Gaussian realization is in much better agreement with the one-loop perturbation theory calculations.   
    \label{fig:init}}
\end{center}
\end{figure}

### 2012-05-23 Introduction.tex
Introduced a naming of the simulation runs for easier reference

Tests on constraints of the redshift step size were performed 
on the Sunnyvale beowulf cluster at the Canadian Institute for Theoretical Astrophysics (CITA).
Each node contains 2 quad core Intel(R) Xeon(R) E5310 @ 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
We generate a set of Gaussian initial conditions at a starting redshift of $z = 100$, 
with a box size equal to $200 h^{-1}\mbox{Mpc}$, we placed
$128^{3}$  particles on a grid with $2^{3}$ more cells, and let them evolve until $z=0$.
Tests on improvements of the force calculations were performed on the same machine, but with 8 times more particles and node, 
and with a box size of $500 h^{-1}\mbox{Mpc}$.
Hereafter we refer to these simulation modes as the CITA128 and CITA256 configurations respectively.

We then test the dynamical range of the simulation, the non-Gaussian initial conditions generator and the run time halo finder algorithm 
with a third simulation configuration, which we denote the  RANGER4000, that also evolves $4000^{3}$ particles 
from $z=100$ to $z=0$ but with a box side of $3.2 h^{-1}\mbox{Gpc}$.
 Much larger runs were produced at various high performance computing centers
 around the world; their description and a comparison of their performance are also described in this paper.

>>>>

Tests on constraints of the redshift step size were performed 
on the Sunnyvale beowulf cluster at the Canadian Institute for Theoretical Astrophysics (CITA).
Each node contains 2 Quad Core Intel(R) Xeon(R) E5310 1.60GHz processors, 4GB of RAM,  a 40 GB disk and 2 gigE network interfaces. 
We generate a set of Gaussian initial conditions at a starting redshift of $z = 100$, 
with a box size equal to $200 h^{-1}\mbox{Mpc}$, we placed
$128^{3}$  particles on a grid with $2^{3}$ more cells, and let them evolve until $z=0$.
Tests on improvements of the force calculations were performed on the same machine, but with 8 times more particles and node, 
and with a box size of $500 h^{-1}\mbox{Mpc}$.
Hereafter we refer to these simulation modes as the CITA128 and CITA256 configurations respectively.

For tests of the accuracy over a large dynamical range of the simulation, of the non-Gaussian initial conditions generator and of the run time halo finder algorithm, 
we used a third simulation configuration that were run at the Texas Advanced Computing Centre (TACC) on Ranger, a SunBlade 
x6420 system with AMD x86\_64 Opteron Quad Core 2.3 GHz `Barcelona' processors and Infiniband networking.
These RANGER4000 simulations evolved $4000^{3}$ particles 
from $z=100$ to $z=0$ with a box side of $3.2 h^{-1}\mbox{Gpc}$
on 4000 cores.

 Much larger runs were produced at various high performance computing centers
 around the world; their description and a comparison of their performance are also described in a later section of this paper.

### 2012-05-23 Systematics.tex
Introduced a naming of the simulation runs for easier reference

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum for a large run ($4000^3$ particles),
which are evolved from a Gaussian initial distribution at $z=100$ down to $z=0$, in a box $3.2 h^{-1}\mbox{Gpc}$ per side.
We observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ is caused by the white noise imposed
in the initial conditions.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The simulation ran on 4000 cores on Ranger.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

The actual force of gravity on the p$^3$m algorithm is presented in Fig. \ref{fig:den_force_ppext0}.
This is a force versus distance plot, where the force is calculated in two steps. 
We first compute the force on each particles of a given distribution as done at every time step.
We then remove one particle, compute the force again on all particles, and record on a file the 
force difference between the two calculations and the distance to the `hole'.

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the p$^3$m algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
  Particles in the same fine cell follow the exact curve, then we observe a large scatter at 
  distance of the order of the fine grid, caused by the NGP interpolation scheme. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

@@

At early stages of the simulation, the density fields is rather homogenous, causing the force of gravity to be
rather weak everywhere. In that case, the size of the redshift jumps is controlled by a limit in the cosmological expansion.
If the expansion jump is too large, the size of the residual errors can become significant, and one can observe, for instance,
a growth of structure that does not match the predictions of  linear theory even at the largest scales.
One therefore needs to choose a maximum step size. In {\small CUBEP3M}, this is controlled by $r_{max}$, which is the fractional step size,
$\mbox{d}a/(a + \mbox{d}a)$ and is set to 0.05 by default.  Generally, a simulation should start at a redshift high enough that
the initial dimensionless power spectrum is well under unity at all scales. This ensures that the Zel'dovich approximation
 holds at the pre cent level at least. A drop of accuracy can occur if one starts the simulation too early, where
 truncation error will be significant at the early time steps.

It is possible to reduce this effect, and thereby improve significantly 
the accuracy of the code, by modifying the value of $r_{max}$, at the cost of increasing the total number of time steps.
Fig. \ref{fig:ra_max} shows a comparison of late time power spectra of density fields that originate from the same initial conditions, and used the same random seeds to control the fine mesh shifts (mentioned above), but the value of $r_{max}$ was modified at each run. 
We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
allows the simulation to recover about 30 per cent of dimensionless power at the turn over scale, in this simulation configuration.
This gain is greatly affected by the choice of initial redshift, the resolution, and the box size, and ideally one would make
test runs in order to optimize a given configuration.  
As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the gain in time is more than a factor of four.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_ra_max.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $500 h^{-1}\mbox{Mpc}$ per side,
  from $256^3$ particles and a starting redshift of $z=200$. The different curves show different values of $r_{max}$. 
  The resources required to run these simulations increase rapidly as $r_{max}$ decreases.    \label{fig:ra_max}}
\end{center}
\end{figure}

>>>>

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum in a RANGER4000 simulation and
observe that the agreement with the non-linear prediction of \cite{Lewis:1999bs} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ are caused by the white noise imposed
in the initial conditions.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a RANGER4000 simulation.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

The actual force of gravity in the p$^3$m algorithm,
as felt by a single particle, is presented in Fig. \ref{fig:den_force_ppext0}.
This shows the force versus distance -- in fine cell units -- a calculation that was performed in a CITA256 realization in two steps: 
1- we compute the force on each particle in a given distribution, as done at every time step.
2- we remove a selected particle, compute the force again on all particles, and record on a file the 
force difference (before  and after the removal) as a function of the distance to the `hole'.

@@

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/densityForce_ppext=0.eps}
  \caption{Gravity force in of the p$^3$m algorithm, versus distance in fine mesh cell units, compared with the exact $1/r^{2}$ law.
  Particles in the same fine cell follow the exact curve, then we observe a large scatter at 
  distance of the order of the fine grid, caused by the NGP interpolation scheme
  This calculation was obtained in a CITA256 realization. 
    \label{fig:den_force_ppext0}}
\end{center}
\end{figure}

@@

At early stages of the simulation, the density fields is rather homogenous, causing the force of gravity to be
rather weak everywhere. In that case, the size of the redshift jumps is controlled by a limit in the cosmological expansion.
If the expansion jump is too large, the size of the residual errors can become significant, and one can observe, for instance,
a growth of structure that does not match the predictions of  linear theory even at the largest scales.
One therefore needs to choose a maximum step size. In {\small CUBEP3M}, this is controlled by $r_{max}$, which is the fractional step size,
$\mbox{d}a/(a + \mbox{d}a)$ and is set to $0.05$ by default.  Generally, a simulation should start at a redshift high enough that
the initial dimensionless power spectrum is well under unity at all scales. This ensures that the Zel'dovich approximation
 holds at the per cent level at least. A drop of accuracy can occur if one starts the simulation too early, where
 truncation error will be significant at the early time steps.

It is possible to reduce this effect, and thereby improve significantly 
the accuracy of the code, by modifying the value of $r_{max}$, at the cost of increasing the total number of time steps.
Fig. \ref{fig:ra_max} shows a comparison of late time power spectra of a series of CITA256 realizations that originate from the same initial conditions, 
and used the same random seeds to control the fine mesh shifts (mentioned above): only the value of $r_{max}$ was modified between each run. 
We observe that the impact is mostly located in the non-linear regime, where decreasing the time step to 0.006 
allows the simulation to recover about 30 per cent of dimensionless power at the turn over scale, in this simulation configuration.
This gain is greatly affected by the choice of initial redshift, the resolution, and the box size, and ideally one would make
test runs in order to optimize a given configuration.  
As expected, the {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}. 
In this test case, reducing further at 0.001 shows only a mild improvement in accuracy, but the gain in time is more than a factor of four.

\begin{figure}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_ra_max.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a series of CITA256 realizations. 
 The starting redshift ofwas raised to $z=200$ to enhance the systematic effect. The different curves show different values of $r_{max}$. 
  The resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.    \label{fig:ra_max}}
\end{center}
\end{figure}
